A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-11-12 inst/manuscript/manuscript.Rmd
finished draft for review

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  This is the abstract.

@@

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus on comparing these approaches for spatial data. We define spatial data as data that incorporates the specific locations of the population units into either the design or estimation process. @de1990model give an early comparison of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @wang2012review; @brus2020statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design and model-based principles (see e.g., @sterba2009alternative, @cicchitelli2012model, @chan2020bayesian for a Bayesian approach).

@@

Instead of estimating fixed but unknown parameters like a mean or total (as in the design-based approach), the goal of model-based inference in the spatial context is often to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data generating process. If we repeatedly generate the response values from the same spatial process and sample, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data. A visual comparison of the design-based and model-based assumptions is provided in Figure \ref{fig:fig1} (@verhoef2002sampling and @brus2020statistical provide similar figures).

@@

We note that we only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. Other methods, such as k-nearest-neighbors [@fix1951discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, could also be used to obtain predictions for a mean or total from spatially correlated responses of a finite population. We choose to use FPBK because it is faster than a Bayesian approach and it was developed with theoretically-based variance estimators of means and totals for spatial data, whereas random forests and k-nearest-neighbors use ad-hoc variance estimators in most cases [@ver2013comparison]; additionally, FBPK outperformed the other methods in most scenarios.

@@

@@

The average bias was nearly zero for all four combinations in all 36 scenarios, so we omit a visualization of those results here. Tables for average bias in all 36 simulation scenarios are provided in the supplementary material.

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and quality in lakes in the contiguous United States, to obtain an interval for mean mercury concentration. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes.

@@

```{r figdata, out.width = "100%", fig.cap = "Population distribution of mercury concentration for 986 lakes in the contiguous United States.", message = FALSE, warning = FALSE}
merc <- ggplot(data = nla_nomiss_both, aes(x = total_hg)) +
  geom_histogram(colour = "black", fill = "white", bins = 15) +
  labs(x = "HG (ng / g)")
# dist <- ggplot(data = nla_nomiss_both, aes(x = RDis_IX)) +
#   geom_histogram(colour = "black", fill = "white", bins = 15) +
#   labs(x = "Lakeshore Disturbance")
merc_map <- ggplot(data = nla_nomiss_both, aes(x = xcoords, y = ycoords)) +
  geom_point(aes(colour = total_hg)) +
  scale_colour_viridis_c() +
  labs(x = "", y = "", colour = "hg") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank()) +
  coord_quickmap() +
  theme_bw(base_size = 14)
# dist_map <- ggplot(data = nla_nomiss_both, aes(x = xcoords, y = ycoords)) +
#   geom_point(aes(colour = RDis_IX)) +
#   scale_colour_viridis_c() +
#   labs(x = "", y = "", colour = "Dist") +
#   theme(axis.text.x = element_blank(),
#         axis.text.y = element_blank(),
#         axis.ticks = element_blank())
library(gridExtra)
# grid.arrange(merc, merc_map, nrow = 2)
merc_map
```

@@

@@

Table \ref{tab:appliedtab} shows the application of a design-based analysis of an IRS sample, a model-based analysis of an IRS sample, a design-based analysis of a GRTS sample, and a model-based analysis of a GRTS sample. For all four analyses, the true realized mean mercury concentration is within the bounds of the 95% intervals. However, we should not generalize the results of this particular realization to any other data set or even to other potential samples of this data set.

<!-- not sure why this is getting cutoff. temporary fix: \pagebreak 
probably because it is an html table -- I used xtable to fix -->

But, we do note a couple of patterns. The design-based IRS analysis shows the largest standard error: a likely reason is that this is the only approach that does not incorporate any spatial information regarding mercury concentration across the contiguous United States. We also see that both approaches using the GRTS sample have a lower standard error than the both approaches using the IRS sample. We would expect this to be the case for most samples because mercury concentration exhibits spatial patterning, so a spatially balanced sample should usually yield a lower standard error.

@@

The design-based and model-based approaches to inference are fundamentally different paradigms by which to samples are selected and data are analyzed. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized value of a random variable. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings are particularly notable: the design decision (GRTS vs IRS) seems much more important than the analysis decision (design-based vs model-based); independent of the analysis approach, there is no reason to prefer IRS over GRTS for spatial data -- GRTS tends to perform at least as well as IRS when there is no spatial correlation and increasingly better than IRS as the strength of spatial correlation increases; the gap in relative efficiency between GRTS-design and GRTS-model widens as the strength of spatial correlation increases; and when the data are skewed, interval coverage for all approaches improves both as the sample size increases and as the strength of correlation increases.

<!-- The design-based approach relies on few assumptions, while the model-based approach relies on rigid distributional ones. -->

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and the response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as likelihood ratio tests or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be heavily considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.

@@

In the supplementary material, we provide summary statistics for all 36 simulation scenarios.

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial data.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference depends on random sampling. In the model-based approach, inference depends on distributional assumptions. In this manuscript, we compare the approaches for finite population spatial data. We first provide relevant background for design-based and model-based approaches to finite population patial data. Then we use a simulation study and an analysis of real mercurcy concentration data to compare them numerically. We find that sampling plans that incorporate spatial locations (spatially balanced samples) perform better than sampling plans ignoring spatial locations (non-spatially balanced samples), regardless of whether design-based or model-based approaches were used to analyze the data. We also find that within sampling plans, model-based approaches tend to outperform design-based approaches, even for skewed data. This gap in performance is small when spatially samples are used but large when non-spatially balanced samples are used.

@@

Though the design-based and model-based approaches apply to statistical inference in a broad sense, we focus on comparing these approaches for spatial data. We define spatial data as data that incorporates the specific locations of the population units into either the design or estimation process. @de1990model give an early comparison of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @wang2012review; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design and model-based principles (see e.g., @sterba2009alternative, @cicchitelli2012model, @chan2020bayesian for a Bayesian approach).

@@

Instead of estimating fixed but unknown parameters like a mean or total (as in the design-based approach), the goal of model-based inference in the spatial context is often to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data generating process. If we repeatedly generate the response values from the same spatial process and sample, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data. A visual comparison of the design-based and model-based assumptions is provided in Figure \ref{fig:fig1} (@verhoef2002sampling and @brus2021statistical provide similar figures).

@@

We note that we only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. Other methods, such as k-nearest-neighbors [@fix1989discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, could also be used to obtain predictions for a mean or total from spatially correlated responses of a finite population. We choose to use FPBK because it is faster than a Bayesian approach and it was developed with theoretically-based variance estimators of means and totals for spatial data, whereas random forests and k-nearest-neighbors use ad-hoc variance estimators in most cases [@ver2013comparison]; additionally, FBPK outperformed the other methods in most scenarios.

@@

The average bias was nearly zero for all four combinations in all 36 scenarios, so we omit a more detailed summary of those results here. Tables for average bias in all 36 simulation scenarios are provided in the supplementary material.

@@

@@

The Environmental Protection Agency (EPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) in the United States to assess the water quality of various bodies of water. We will use the 2012 National Lakes Assessment (NLA), which measures various aspects of lake health and quality in lakes in the contiguous United States, to study mercury concentration. Although we know the true mean mercury concentration values for the 986 lakes from the 2012 NLA, we will explore whether or not we obtain an adequately precise estimate for the realized mean mercury concentration if we sample only 100 of the 986 lakes.

@@

@@

```{r figdata, out.width = "49%", fig.cap = "Population distribution of mercury concentration (hg) for 986 lakes in the contiguous United States in a spatial layout (left) and a histogram (right).", message = FALSE, warning = FALSE, fig.show='hold', fig.align='center'}
merc <- ggplot(data = nla_nomiss_both, aes(x = total_hg)) +
  geom_histogram(colour = "black", fill = "white", bins = 15) +
  labs(x = "HG (ng / g)")
# dist <- ggplot(data = nla_nomiss_both, aes(x = RDis_IX)) +
#   geom_histogram(colour = "black", fill = "white", bins = 15) +
#   labs(x = "Lakeshore Disturbance")
merc_map <- ggplot(data = nla_nomiss_both, aes(x = xcoords, y = ycoords)) +
  geom_point(aes(colour = total_hg)) +
  scale_colour_viridis_c() +
  labs(x = "", y = "", colour = "Hg") +
  lims(x = c(0, 6500)) +
  coord_quickmap() +
  theme_bw(base_size = 20) + 
  theme(axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      legend.position = c(0.85, 0.5),
      legend.key.size = unit(0.75, "cm"))

merc_hist <- ggplot(data = nla_nomiss_both, aes(x = total_hg)) +
  geom_histogram(bins = 50) + 
  labs(x = "Hg", y = "Count") +
  theme_bw(base_size = 20)

merc_map
merc_hist
```

@@

We selected a single IRS sample and a single GRTS sample and estimated the mean mercury concentration and its standard error using using design-based and model-based approaches; Table \ref{tab:appliedtab} shows the results. For all four sampling-analysis combinations, the true realized mean mercury concentration is within the bounds of the 95% intervals. However, we should not generalize these results to any other data or even to other samples from these data. But, we do note a couple of patterns. The design-based IRS analysis shows the largest standard error: a likely reason is that this is the only approach that does not incorporate any spatial information regarding mercury concentration across the contiguous United States. We also see that both approaches using the GRTS sample have a lower standard error than the both approaches using the IRS sample. We would expect this to be the case for most samples because mercury concentration exhibits spatial patterning, so a spatially balanced sample should usually yield a lower standard error.

To better understand the dependence structure in mercury concentration, the empirical semivariogram and corresponding fit of the model-based approaches can be visualized. The empirical semivariogram quantifies the halved squared differences (semivariance) among response values at different distances apart. If a process exhibits strong spatial dependence, the empirical semivariogram will have small values at small distances and large values at large distances. Figure \ref{fig:figsv} shows the empirical semivariogram for GRTS-Model, displaying the average semivariance for several distances.  Overlain onto Figure \ref{fig:figsv} is the estimated semivariance obtained using the covariance parmaeters from the REML fit of GRTS-Model. Figure \ref{fig:figsv} provides evidence that there is strong correlation in mercury concentration among the sites.

```{r figsv, out.width = "85%", fig.align='center', fig.cap = "The empirical semivariogram (black circles) of mercury concentration against the REML fit using the estimated covariance parameters (black line) from GRTS-Model."}
plot(mod_grts)
```

@@

The design-based and model-based approaches to inference are fundamentally different paradigms by which samples are selected and data are analyzed. The design-based approach incorporates randomness through sampling to estimate a population parameter. The model-based approach incorporates randomness through distributional assumptions to predict the realized values of a random process. Though these approaches have often been compared in the literature both from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling. Aside from the theoretical differences described, a few analytical findings from the simulation study are particularly notable. First, the sampling decision (GRTS vs IRS) is most important when using a design-based analysis. Though GRTS-Model still outperformed IRS-Model, the model-based analysis mitigated much of the inefficiency of the IRS sample. Second, independent of the analysis approach, there is no reason to use IRS over GRTS for sampling spatial data, as GRTS-Design and GRTS-Model generally performed at least as well as their IRS counterparts when there was no spatial correlation and noticeably better than there IRS counterparts when there was spatial correlation. Third, The stronger the spatial correlation, the larger the gap in rMS(P)E between IRS-Design and the other sampling-analysis combinations. Fourth and finally, interval coverage for the normal response was very close to 95% for all sample sizes, while interval coverage for the lognormal response was not very close to 95% until $n = 200$.

There are several benefits and drawbacks of the design-based and model-based approaches for spatial data, some of which we have not yet discussed but are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based estimation of covariance parameters can be computationally burdensome, especially for likelihood-based methods such as REML that rely on inverting a covariance matrix. The design-based approach also more naturally handles binary data, free from the more complicated logistic regression formulation commonly used to handle binary data in a model-based approach. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and the response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the process of study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed sites. Model-based approaches can also compute site-by-site predictions at unobserved locations and use them to construct informative visualizations. The benefits and drawbacks of both approaches, alongside our theoretical and analytical comparisons, should be seriously considered when choosing among them. This is especially true from an analysis perspective, as we found that using a spatially balanced sampling algorithm benefits both design-based and model-based analyses.

@@

In the supplementary material, we provide tables presenting summary statistics for all 36 simulation scenarios.
